---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

1. Code for reading in the dataset and/or processing the data

```{r unzip, cache=TRUE}
unzip("repdata_activity.zip",exdir = "output")
```

Reading the data into activity data frame.
```{r reading, cache=TRUE}
activity <- read.csv("output/activity.csv", stringsAsFactors=FALSE)
str(activity)
summary(activity)
```

Convert date to POSIXct class using lubridate package and convert interval to hour:minute format
```{r date_conversion, message=FALSE}
library(lubridate)
activity$date <- ymd(activity$date)
str(activity)
```

2. Histogram of the total number of steps taken each day

```{r daily_total,message=FALSE}
require(dplyr)
total_day <- activity %>% group_by(date) %>%summarise(total_steps=sum(steps,na.rm=TRUE),na=mean(is.na(steps))) %>% print
```
Visualise the total number of steps taken per day as a barplot

```{r barplot,fig.width=12}
barplot(height = total_day$total_steps,names.arg=total_day$date,cex.names=0.68,las=3,col="grey")
abline(h=median(total_day$total_steps), col="black")
abline(h=mean(total_day$total_steps), col="blue")
text(x = 0,y=median(total_day$total_steps),pos=3,labels = "median")
text(x = 0,y=mean(total_day$total_steps),pos=1,labels = "mean")
```

3. Make a histogram of the total number of steps taken each day

```{r histogram}
total_day <- filter(total_day, na < 1)
hist(total_day$total_steps,col="grey",breaks=20,main="Total steps per day",xlab="Steps per day")
abline(v=median(total_day$total_steps),lty=3, lwd=2, col="black")
legend(legend="median","topright",lty=3,lwd=2,bty = "n")
```

4. Calculate and report the mean and median of the total number of steps taken per day

```{r summary}
mean_steps <- mean(total_day$total_steps,na.rm=TRUE)
median_steps <- median(total_day$total_steps,na.rm=TRUE)
```
Mean and median of the total number of steps taken per day are `r round(mean_steps,2)` steps and `r median_steps` steps, respectively.

5. What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)  

```{r daily, fig.width=12}
library(dplyr,quietly = TRUE)
daily_patterns <- activity %>% group_by(interval) %>% summarise(average=mean(steps,na.rm=TRUE))
plot(x = 1:nrow(daily_patterns),y = daily_patterns$average,type = "l",
     col = "black", xaxt = "n",xlab="Intervals", 
     ylab = "Average for given interval across all days")
axis(1,labels=daily_patterns$interval[seq(1,288,12)],
     at = seq_along(daily_patterns$interval)[seq(1,288,12)])

```
```{r daily2}
max_numb_steps_interval <- filter(daily_patterns,average==max(average))
```

Interval **"`r max_numb_steps_interval$interval`"** contains on average the maximum number of steps (**`r round(max_numb_steps_interval$average,2)` steps**).


Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing}
na_number <- sum(is.na(activity$steps))
na_number
percentage_na <- mean(is.na(activity$steps))
percentage_na
```
Total number of missing values in the dataset amounts to **`r na_number` ** (what is **`r round(percentage_na*100,1)`** % of total observations).

Devise a strategy for filling in all of the missing values in the dataset

As the number of missing values in this dataset is fairly large, we cannot be sure if there is no bias introduced by missing values. Therefore we impute missing values based on average number of steps in particular 5-minutes interval. 

6. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r na_imputing, cache=TRUE}
without_NAs <- numeric(nrow(activity))
for (i in 1:nrow(activity))
{
        if (is.na(activity[i,"steps"])==TRUE)
            {
                    without_NAs[i]<-filter(daily_patterns,interval==activity[i,"interval"]) %>% select(average)
            } 
        else
            {
                    without_NAs[i]<-activity[i,"steps"]
            }
                    
}
activity_without_NAs<-mutate(activity,steps_no_NAs=without_NAs)
head(activity_without_NAs)
```

Below code is just to verify if process of imputing missing values correctly preserved original values (lines with no NAs)   
```{r test_NAs}
check <- filter(activity_without_NAs,!is.na(steps)) %>% mutate(ok = (steps==steps_no_NAs))
mean(check$ok)
```

7. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day

```{r histogram_no_NAs}
total_day_noNAs <- activity_without_NAs %>% mutate(steps_no_NAs=as.numeric(steps_no_NAs)) %>% group_by(date) %>% summarise(total_steps=sum(steps_no_NAs))
hist(total_day_noNAs$total_steps,col="yellow",breaks=20,main="Total steps per day",xlab="Steps per day")
abline(v=median(total_day$total_steps),lty=3, lwd=2, col="black")
legend(legend="median","topright",lty=3,lwd=2,bty = "n")
```

```{r summary_no_NAs}
summary(total_day_noNAs$total_steps)
```

Imputing missing values, mean of the total number of steps taken per day  increased while median decreased,compared to estimates from the first part (ingoring missing values). Imputing missing data resulted in increase of total daily number of steps (instead of each NAs we have average that is always >=0)

```

8. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day

```{r weekday}
library(lubridate)
is_weekday <-function(date){
        if(wday(date)%in%c(1,7)) result<-"weekend"
        else
                result<-"weekday"
        result
}
activity_without_NAs <- mutate(activity_without_NAs,date=ymd(date)) %>% mutate(day=sapply(date,is_weekday))
table(activity_without_NAs$day)
```

9. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)


```{r weekend_comparison, cache=TRUE, fig.width=10}
library(ggplot2)
daily_patterns <- activity_without_NAs %>% mutate(day=factor(day,levels=c("weekend","weekday")),steps_no_NAs=as.numeric(steps_no_NAs)) %>% group_by(interval,day) %>% summarise(average=mean(steps_no_NAs))
qplot(interval,average,data=daily_patterns,geom="line",facets=day~.)
```
